---
title: "COVID-19 Vaccines: Clinical Trial Participation vs Real World Use"
subtitle: "Data Wrangling and Exploratory Analysis"
author: "Clare Gibson"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: paper
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
# Load knitr package
library(knitr)

# Knitr Options
opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	fig.align = 'center'
)
```

# Introduction
This analysis explores relationships between COVID-19 vaccine trial participation and real world vaccination doses administered.

## Data Sources
The data for this analysis comes from the following sources:

* [AACT Database](https://aact.ctti-clinicaltrials.org/): an open-source relational database that contains all data points for every study registered on ClinicalTrials.gov.
* [Bloomberg Graphics Covid Vaccine Tracker Data](https://github.com/BloombergGraphics/covid-vaccine-tracker-data): open dataset covering all US states and a growing number of countries internationally.

# Data Wrangling
First I load the packages I need for this analysis.
```{r libraries}
library(RPostgreSQL)    # for connecting to a PostgreSQL DB
library(rstudioapi)     # for hiding credentials
library(tidyverse)      # for data wrangling
```

Next I need to establish a connection to the AACT database.
```{r connect-to-aact}
# Load the database driver
#drv <- dbDriver("PostgreSQL")

# Create the connection
#con <- dbConnect(drv,
#                 dbname="aact",
#                 host="aact-db.ctti-clinicaltrials.org",
#                 port=5432,
#                 user=askForPassword("Database user"),
#                 password=askForPassword("Database password"))
```

# Data Wrangling