---
title: "Exploring Gender Balance in Clinical Trial Participation"
subtitle: "Data Wrangling and Exploratory Analysis"
author: "Clare Gibson"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: paper
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
# Load knitr package
library(knitr)

# Knitr Options
opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	fig.align = 'center'
)
```

# Introduction
Using data from [ClinicalTrials.gov](https://clinicaltrials.gov/), this paper attempts to evaluate the distribution of male and female participants in clinical research, across different therapeutic areas. The data I am using for this analysis comes from the [AACT Database](https://aact.ctti-clinicaltrials.org/), an open-source relational database that contains all data points for every study registered on ClinicalTrials.gov.

## Questions
I plan to focus my analysis on the following questions:

* What is the ratio of male to female participants in completed interventional trials that included sites in the US?

## Set-up
First I load the packages I need for this analysis.
```{r libraries}
library(RPostgreSQL)    # for connecting to a PostgreSQL DB
library(rstudioapi)     # for hiding credentials
library(tidyverse)      # for data wrangling
```

Next I need to establish a connection to the AACT database.
```{r connect-to-aact}
# Load the database driver
drv <- dbDriver("PostgreSQL")

# Create the connection
con <- dbConnect(drv,
                 dbname="aact",
                 host="aact-db.ctti-clinicaltrials.org",
                 port=5432,
                 user=askForPassword("Database user"),
                 password=askForPassword("Database password"))
```

# Data Wrangling
Now I can query the database for data on all trials that include the US as a country. The [data dictionary](ref/aact_data_definitions.xlsx) for the AACT database helps me to navigate the tables and columns.

```{r query-the-db}
# Write the SQL query
query <- 
  "SELECT
     s.nct_id,
     s.start_date,
     s.official_title,
     s.phase,
     d.name,
     b.classification,
     b.category,
     b.param_type,
     b.param_value_num
   FROM studies s
   INNER JOIN conditions d
   ON s.nct_id = d.nct_id
   INNER JOIN baseline_measurements b
   ON s.nct_id = b.nct_id
   INNER JOIN result_groups r
   ON s.nct_id = r.nct_id
   AND b.ctgov_group_code = r.ctgov_group_code
   WHERE s.nct_id IN (
     SELECT DISTINCT nct_id
     FROM countries c
     WHERE name = 'United States'
   )
   AND s.overall_status = 'Completed'
   AND b.title = 'Sex: Female, Male'
   AND b.units = 'Participants'
   AND b.param_value_num IS NOT NULL
   AND r.title = 'Total'"

# Save the results of the query to a df
df_raw <- dbGetQuery(con, query)
```

Now I can view the first few lines of the dataframe from this query.
```{r df-head}
# Display the first 5 rows of the df
df_raw %>% 
  head(5) %>% 
  kable()
```

I will aggregate the results so that I report the total number of male and female participants by study and condition.
```{r aggregate-df}
# Group by study and condition and sum the male/female participants
df_clean <- df_raw %>% 
  group_by(nct_id,
           name,
           category) %>% 
  mutate(participant_count = sum(param_value_num)) %>% 
  select(-param_value_num) %>% 
  distinct() %>% 
  ungroup()
```

Now I need to pivot the data to place female and male participants in their own columns and calculate the total participation by summing the two values.
```{r df-pivot}
# Pivot the gender counts into their own columns
df_clean <- df_clean %>% 
  mutate(category=str_to_lower(category)) %>% 
  pivot_wider(names_from=category,
              values_from=participant_count) %>% 
  mutate(total = female + male,
         female_pct = female / total,
         male_pct = male / total)
```


# Export
Finally I can export my query to a CSV file for further analysis in Tableau.
```{r export-csv}
# Write a CSV output of the df
write_csv(x=df_clean,
          file="data-out/gender.csv",
          na="")
```

